|

### 2021-09-15 paper.Rmd
bullet points for theory

<!-- % emotion in political science -->
In this study, I follow recent work in political science inspired by moral psychology. Most relevant for this study, @clifford2019emotional finds that a persuasive emotional frame loading on emotions of anger and disgust increases the perception of an issue as being one of moral conviction [@skitka2020psychology; @wisneski2017moralization]. 
Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude].

@@

<!-- % Final Punchline on broader relevance -->
In the view of many scientists [@roberts2018nobel, planck2019scientists], genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations.

# Theory

@@

@@

Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information.

The effects on related opinions are much less clear.
Some studies find effects, some not.

Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected.

## Emotions, political learning, moralization {-}

@@

Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.

Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.

Research has shown that... anger... anxiety... enthusiasms...

A recently growing interest is in disgust.
Connection to morality, relevant for immigration, homeless ...

A study has shown that moral shock works to moralize attitudes towards GM food.

While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].

However, the main part of the material used there can be classified as misinformation
GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So

@@

## GMO - a case for emotional reactions and misinformation {-}

<!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
The topic of GM feed is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, hasell2020differential, carnahan2019processing]. This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans]. GM food is also a topic on which there are misperceptions, possibly due to disinformation. Of course, there is legitimate criticism on the use of genetic modification. However, studies show that extreme opponents of GM food know the least about it [@fernbach2019extreme], and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].

@@

An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
Because all shown information has in a similar format been in the media, there is no deception taking place.
The experimental treatment intends to evoke emotions, but I consider them within reason. There are two main reasons for that. First, the material including the used imagery has been used in a previously published study @[clifford2019emotional]. Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents. Because this approval has been granted, this study did not require further approval by an ethics committee.

@@

The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon. This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information @[lewandowsky2020debunking]. As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format. By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food. This is done in a further condition, where the corrective information is embedded in a counter-frame, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.

Assignment is done by simple random assignment into equal sized groups. Table gives an overview over the nine resulting experimental groups. The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned). For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' @[kuklinski2000misinformation, p.805].

@@

To provide a broad measure of policy support, I use a measure by @[diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".

I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" @[skitka2020psychology]. This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them. Both items will be combined into a simple additive index.

I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food. Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds @[skitka2018attitude]. The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".

@@

Table \@ref(tab:descriptives) shows descriptive statistics of the outcome variables in the control group, which does not get any informational treatment and thus serves as the reference to investigate information effects. On the scale from 0 to 4, anger as an emotional reaction to th issue of GM food was expressed with a mean of 1.7 moderately below the midpoint. Anxiety was above the midpoint with a mean of 2.4 whereas disgust was realtively low with a mean of 1.3, and sadness again moderatly below the midpoint with 1.7. These figures indicate that respondents' negative reactions to GM food stem from concerns rather than disgust.

@@

In the appendix, I report the regression tables, in the following I will visualize the estimated regression coefficients for the different treatment conditions.

@@

## Counter-frame {-}

@@

# Conclusion

@@

* is it really emotions or "just" the pictures?

@@

* When misinformation is emotionally charged, fact-checkers may have to think about counter-frames to eliminate the misinformation effect.
* a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system @siegrist2020consumer.  genetic engineering provides avenues with large potential benefits @roberts2018nobel, @planck2019scientists, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
* Similar effects possible on other issues where misinformation loads on strong emotions, e.g. health interventions such as vaccinations.

>>>>

<!-- % emotion in political science -->
In this study, I follow recent work in political science inspired by moral psychology. Most relevant for this study, @clifford2019emotional finds that a persuasive emotional frame loading on emotions of anger and disgust increases the perception of an issue as being one of moral conviction [@skitka2020psychology; @wisneski2017moralization]. 
Moral conviction, i.e. the conviction that the handling of an issue touches morality and is about fundamental right or wrong, can be seen as a dimension of attitude strength [@skitka2018attitude]. 
While it is often seen as a binary categorization whether or not an issue is a moral one, there is quite some variation in the extent of moralization within issues [@ryan2014reconsidering] with distinct behavioral consequences, such as an absolute unwillingness to compromise [@ryan2017no] or a tendency to view that issue in absolutist term and to be less likely to be open for information about consequences [@ryan2019actions].

@@

<!-- % Final Punchline on broader relevance -->
In the view of many scientists [@roberts2018nobel, @planck2019scientists], genetic engineering provides avenues with large potential benefits, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations.

\newpage
# Theory

@@

*t.b.d.*

* Definition and distinction between misinformation and disinformation.
* Some worries about backfire effects, but usually, corrective information works, and guidelines for best practice corrections have been collected.
* Emotions may contribute to the spread of false information.
* Relatively little evidence on the role of emotions for misinformation and corrective information.

@@

<!-- Factual beliefs are affected in the direction of presented evidence for both (a) misinformation and (b) corrective information. -->

<!-- The effects on related opinions are much less clear. -->
<!-- Some studies find effects, some not. -->

<!-- Affective responses are at least implicitly discussed in this literature, but discrete emotions are widely neglected. -->

## Emotional framing and moralization {-}

*t.b.d.*

* Emotions feature prominently in political communication.
* Framing.
* Some known effects on information processing.
* Findings from moral psychology on moralization through disgust.
* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.

@@

<!-- Emotions certainly play an important role in politics [@redlawsk2006feeling], and approaches such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.  -->

<!-- Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored. -->

<!-- Research has shown that... anger... anxiety... enthusiasms... -->

<!-- A recently growing interest is in disgust. -->
<!-- Connection to morality, relevant for immigration, homeless ... -->

<!-- A study has shown that moral shock works to moralize attitudes towards GM food. -->

<!-- However, the main part of the material used there can be classified as misinformation -->
<!-- GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So -->

@@

## GM food - a case for emotional reactions and misinformation {-}

<!-- % explain the relevance of GMO as an issue, also in realtion to moralization -->
The topic of GM food is particularly suited for this study because it is an issue where science and public opinion diverge considerably, implying a relevant role for corrective information [@diamond2020does, @hasell2020differential, @carnahan2019processing]. This is especially true within the German population, which is where this study draws its sample from [@gaskell2010europeans]. GM food is also a topic on which there are misperceptions, possibly due to disinformation. Of course, there is legitimate criticism on the use of genetic modification. However, studies show that extreme opponents of GM food know the least about it [@fernbach2019extreme], and a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system [@siegrist2020consumer].

@@

An extensive debriefing is included at the end of the study for all respondents. Specifically, all original sources of information are revealed.
Because all shown information has in a similar format been in the media, there is no deception taking place.
The experimental treatment intends to evoke emotions, but I consider them within reason. There are two main reasons for that. First, the material including the used imagery has been used in a previously published study [@clifford2019emotional]. Second, the central part of the visual material has been used by actual reporting on the issue in mainstream media^[www.welt.de/newsticker/news1/article109594074/Bundesinstitut-Genmais-Schaeden-nicht-ausreichend-belegt].
The study was approved by the team operating the GIP, with the goal of designing the study in a way that it is an appropriately pleasant experience for respondents. Because this approval has been granted, this study did not require further approval by an ethics committee.

@@

The second type of information reports the accurate scientific consensus that GM food is generally safe to eat, that a study showing cancer growth in rats due to GM feed has been retracted, and that genetically modifying plants for the food industry is not a new phenomenon. This piece of information is a correction to the misinformation in that it reflects best available scientific evidence and it is constructed to conform to best practice standards of debunking information [@lewandowsky2020debunking]. As a neutral correction, this piece of information corrects the misinformation about concrete health risks from GM food in a text-only format. By negating the risks, it still stays within the overall framing of GM food as possibly risky, and does not add specific further information about benefits from GM food. This is done in a further condition, where the corrective information is embedded in a counter-frame, emphasizing positive aspects of GM food, such as its potential to contribute to more sustainable agriculture by reducing pesticides, and its potential to fight malnutrition.

Assignment is done by simple random assignment into equal sized groups. Table gives an overview over the nine resulting experimental groups. The misinformation (if any is assigned) is always shown first, immediately followed by the corrective information (if any is assigned). For respondents, the manipulation is thus designed in a way that both types of information ``hits them between the eyes'' [@kuklinski2000misinformation, p.805].

@@

To provide a broad measure of policy support, I use a measure by [@diamond2020does] that asks respondents how high a policy priority it should be to restrict the use of GM food on a 7-point scale from "no priority" to "highest priority".

I measure moral conviction with two standard items from the literature, asking respondents whether their evaluation of GM food is connected to moral convictions and about fundamental right and wrong, respectively, on a 5-point scale from "not at all" to "very much" [@skitka2020psychology]. This is a direct measure of a "meta-cognition", building on the assumption that respondents know which issues are about moral conviction for them. Both items will be combined into a simple additive index.

I measure distinct emotions by directly asking respondents how well a given feeling represents how they feel toward GM food. Asking about their feelings toward GM food makes this a measure of integral, not incidental emotion, in line with expectation about to which kind of emotion moralization responds [@skitka2018attitude]. The distinct negative emotions I measure are disgust, anger, anxiety, and sadness, each on a 5-point scale from "not at all" to "extremely".

@@

Table \@ref(tab:descriptives) shows descriptive statistics of the outcome variables in the control group, which does not get any informational treatment and thus serves as the reference to investigate information effects. On the scale from 0 to 4, anger as an emotional reaction to th issue of GM food was expressed with a mean of 1.7 moderately below the midpoint. Anxiety was above the midpoint with a mean of 2.4 whereas disgust was relatively low with a mean of 1.3, and sadness again moderately below the midpoint with 1.7. These figures indicate that respondents' negative reactions to GM food stem from concerns rather than disgust.

@@

In the appendix, I report the regression tables, in the following I will visualize the estimated regression coefficients for the different treatment conditions. Without any controls, the coefficients refer to the difference in the mean of the outcome between the control and the respective treatment group. ^[The significance of the coefficients is equivalent to running separate t-tests between the group means, which raises the issue of adjusting for multiple testing. There are different approaches from different traditions to that. When conducting many independent t-tests, it seems straight-forward to adjust for multiple testing, although the general advise to always adjust for multiple testing is not warranted [@althouse2016adjust]. Another approach is to conduct ANOVA, which is an omnibus test, with the null-hypothesis of no effect of the treatment conditions together versus the control. To check for individual differences between experimental groups, post-hoc tests can be conducted with respective appreciation of multiple testing adjustments. This is appropriate if one has no specific hypotheses to begin with. If one has these hypotheses, however, one can use ANOVA with planned contrasts (e.g. [@field2012discovering]), which is then again equivalent to the results of a regression with dummy variables coded according to the specified contrasts. When I use multiple outcomes in my pre-registered hypotheses, the expectation is not that any of the outcomes shows an effect (in which case adjustment for multiple testing should be done), but rather a joint pattern of effects.]

@@

## Counterframe correction {-}

@@

\newpage
# Conclusion

*t.b.d.*

@@

* No distinction between emotional framing and visual cues.

@@

* When misinformation is emotionally charged, fact-checkers may have to think about counterframes to eliminate the misinformation effect.
* a lack of acceptance of novel food technologies against scientific evidence can be seen as as a threat for the development of a more resilient food system @siegrist2020consumer.  genetic engineering provides avenues with large potential benefits @roberts2018nobel, @planck2019scientists, which may be impeded by public resistance possibly originating from misleading claims easily disseminated through online media.
* Similar effects possible on other issues where misinformation loads on strong emotions, e.g. health interventions such as vaccinations.
